---
title: "Analysis effects Extreme Weather USA"
author: "Ben Kooi"
date: "3-1-2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
This analysis of the effects of extreme weather, gives answers to the these two questions:

1. Across the United States, which types of events are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

The used dataset is the StormData-dataset of the National Oceanic and Atmospheric Administration and can be downloaded [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).This dataset contains data from 1950 until november 2011

This analysis shows that the **most harmful** weathertype are **tornado's**. Tornado's have the most fatalities and the most injuries of all weathertypes.
The weathertype with the **most economic property damage** are **floods**.


## Data processing
This section describes the steps that are taken for processing the raw data. These steps are:

1. Load the packages we need for this analysis;
2. download the raw data from the National Weather Service;
3. read the dataset into a datframe called 'stormdata_raw';
4. examine the structure of the raw dataset;
5. create a new dataset 'stormdata_processed' which will be the analysis-dataset;
6. examine the structure, summary and brief

### load the needed packages
``` {r , cache=TRUE}
#load packages
packages <- c("data.table","dplyr", "ggplot2", "knitr")
sapply(packages, require, character.only=TRUE, quietly=TRUE)
```

### download file and read the datafile
``` {r , cache=TRUE}
path_files <- getwd()
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, "StormData.csv.bz2")
stormdata_raw <- read.csv("StormData.csv.bz2")
str(stormdata_raw)
```

### create the analysis-dataset
For answering the two main questions, we only need the variables:

* EVTYPE       = type of weather-event;
* FATALITIES   = amount of persons who died at a particular extreme weather-event;
* INJURIES     = amount of persons who were injured at a particular extreme weather-event;
* PROPDMG      = amount of economic property damage (dollars) at a particular extreme weather-event;
* PROPDMGEXP   = the exponent value of the economic property damage.

For creating the right dataset we have to mutate EVTYPE to a factor-value and calculate the property damage by it's exponent for preparing to summarize. The exponent PROPDMGEXP itself is deleted.

``` {r , cache=TRUE}
stormdata_processed <- stormdata_raw %>%
                        select(EVTYPE,
                               FATALITIES,
                               INJURIES,
                               PROPDMG,
                               PROPDMGEXP) %>%
                        mutate(EVTYPE=as.factor(EVTYPE),
                               PROPDMG=PROPDMG*ifelse(PROPDMGEXP=='K',1000,
                                                      ifelse(PROPDMGEXP=='M',1000000,
                                                             ifelse(PROPDMGEXP=='B',1000000000,PROPDMG))),
                               PROPDMGEXP=NULL,
                               ) %>%
                        group_by(EVTYPE) %>%
                        summarize(FATALITIES_TOTAL=sum(FATALITIES),
                                  INJURIES_TOTAL=sum(INJURIES),
                                  PROPDMG_TOTAL=sum(PROPDMG))
```

### examine the analysis-dataset and determine if the dataset is right
The following code tells us if the processed dataset is appropiate to answer the questions.

* the EVTYPE-variable is now a factor-value;
* the PROPDMG is calculated with it's exponent for prparing to summarize;
* the totals of FATALITIES, INJURIES and PROPDMG are calculated by weathertype.
* a table with the first 6 rows is displayed.

``` {r , cache=TRUE}
str(stormdata_processed)
kable(head(stormdata_processed),caption="First 6 rows of the processed dataset")
```

```{r , cache=TRUE}
summary(stormdata_processed)
```

The summary of the processed dataset shows:

* a maximum value of `r max(stormdata_processed$FATALITIES_TOTAL)` fatalities;
* a maximum value of `r max(stormdata_processed$INJURIES_TOTAL)` injuries;
* a maximum value of `r max(stormdata_processed$PROPDMG_TOTAL)` economic property damage.

The maximum values of FATALITIES_TOTAL, INJURIES_TOTAL and PROPDMG_TOTAL in the summary of the dataset are used for checking the plots

## Results
Three barplots are created showing that tornado's are the most harmful- and floods the most damaging weather events.
In preparation for the plots the next actions are taken:

1. create a dataframe for analyzing the most harmful weather events. The data is sorted descending by total fatalities and injuries;
2. create a dataframe for analyzing the most damaging weather events. The data is sorted descending by total damage properties;
3. the top 6 rows of both datasets are shown.
```{r , cache=TRUE}
stormdata_harmful<- stormdata_processed %>% arrange(desc(FATALITIES_TOTAL),desc(INJURIES_TOTAL),desc(PROPDMG_TOTAL))
stormdata_damage<- stormdata_processed %>% arrange(desc(PROPDMG_TOTAL),desc(FATALITIES_TOTAL),desc(INJURIES_TOTAL))
kable(head(stormdata_harmful), caption="Top 6 most harmfull weather events")
kable(head(stormdata_damage),caption="Top 6 most economic property damaging weather events")
```

### Final barplots
The final three barplots are created by plotting the top 6 rows of the datasets. The bars are in descending order of the total values. The first bar of each barplot is showing:

* the most fatalities by type of weather event;
* the most injuries by type of weather event;
* the most damaging by type of weather event;

As you can see the values of the 1st bar in the plots are corresponding the maximum values in the summary of the processed dataset. So we can assume that the plots are well created.

```{r ,cache=TRUE}
ggplot(head(stormdata_harmful), aes(reorder(x=EVTYPE,-FATALITIES_TOTAL), y=FATALITIES_TOTAL)) + 
  geom_bar(stat = "identity",fill="red", colour="red4") +
  labs(title = "Fatalities by Weather events", subtitle="from 1950 until 2011", caption = "Data source: National Oceanic and Atmospheric Administration", x="Type of Weather events", y="Total fatalities")
```


```{r ,cache=TRUE}
ggplot(head(stormdata_harmful), aes(reorder(x=EVTYPE,-INJURIES_TOTAL), y=INJURIES_TOTAL)) + 
  geom_bar(stat = "identity",fill="steelblue1", colour="steelblue4") +
  labs(title = "Injuries by Weather events", subtitle="from 1950 until 2011", caption = "Data source: National Oceanic and Atmospheric Administration", x="Type of Weather event", y="Total injuries")
```

```{r ,cache=TRUE}
ggplot(head(stormdata_damage), aes(reorder(x=EVTYPE,-PROPDMG_TOTAL), y=PROPDMG_TOTAL)) + 
  geom_bar(stat = "identity",fill="hotpink", colour="hotpink4") +
  labs(title = "Economic property damage by Weather events", subtitle="from 1950 until 2011", caption = "Data source: National Oceanic and Atmospheric Administration", x="Type of Weather event", y="Total damage (dollars)")
```

As already stated in the synopsis the plots are showing that  the **most harmful** weathertype are **tornado's** and the weathertype with the **most economic property damage** are **floods**.